---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# silaR

<!-- badges: start -->
[![R-CMD-check](https://github.com/bilgelm/silaR/actions/workflows/check-R-and-matlab.yaml/badge.svg)](https://github.com/bilgelm/silaR/actions/workflows/check-R-and-matlab.yaml)
[![pre-commit.ci status](https://results.pre-commit.ci/badge/github/bilgelm/silaR/main.svg)](https://results.pre-commit.ci/latest/github/bilgelm/silaR/main)
<!-- badges: end -->

silaR is an R implementation of Sampled Iterative Local Approximation (SILA).

## Installation

You can install the development version of silaR from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("bilgelm/silaR")
```

## Example

As a simple example, we illustrate how to run ILLA without smoothing in a
dataset consisting of two subjects, and then how to localize these two 
subjects on the estimated trajectory.

```{r example}
library(silaR)

set.seed(42)

# generate a simple small dataset
df <- tibble::tibble(
  subid = c(
    1, 1, 1,
    2, 2, 2, 2, 2, 2
  ),
  age = c(
    seq(from = 50, to = 70, length.out = 3),
    seq(from = 50, to = 70, length.out = 6)
  ),
  val = c(
    2, 4, 6,
    2, 4, 6, 8, 10, 12
  ) + stats::rnorm(3 + 6, mean = 0, sd = .1)
)

# run ILLA without smoothing
res <- illa(df, dt = 2, val0 = 2, maxi = 100, skern = 0)
```

```{r exampleplot, fig.cap=c("Figure 1. Observed longitudinal biomarker data versus age.", "Figure 2. Estimated biomarker data as a function of aligned time."), out.width="50%"}
library(ggplot2)

df$Subject <- factor(df$subid)

ggplot(df, aes(x = age, y = val, group = Subject, color = Subject)) +
  geom_point() +
  geom_line() +
  xlim(c(50, 70)) +
  ylim(c(2, 13)) +
  xlab("Age (years)") +
  ylab("Biomarker value") +
  theme_bw(base_size = 20) +
  theme(legend.position = "inside", legend.position.inside = c(.15, .8))

adtime_lims <- c(-12, 25)
ggplot(res$tout, aes(x = adtime, y = val)) +
  geom_line(linewidth = 2, color = "gray", linetype = "dashed") +
  xlim(adtime_lims) +
  ylim(c(2, 13)) +
  xlab("Adjusted time (years)") +
  ylab("Biomarker value") +
  theme_bw(base_size = 20)
```

We can now estimate the time shift for each subject by projecting their data
onto the estimated trajectory.

```{r silaestimateplot, fig.cap=c("Figure 3a. Alignment using last visit. Observed value at last visit intersects the estimated (extrapolated) trajectory shown in gray.", "Figure 3b. Alignment using first visit. Observed value at first visit intersects the estimated (extrapolated) trajectory shown in gray.", "Figure 3c. Alignment using all visits. Each subject is shifted such that the sum of squared errors with the estimated (extrapolated) trajectory (shown in gray) is minimized."), out.width="33%"}
# run SILA estimate
res_estimate_last <- sila_estimate(res$tout, df)

ggplot(res_estimate_last, aes(x = estdtt0, y = val, color = Subject)) +
  geom_point() +
  geom_line(aes(group = Subject)) +
  geom_line(
    data = res$tout,
    aes(x = adtime, y = val),
    linewidth = 2,
    color = "gray",
    linetype = "dashed"
  ) +
  xlim(adtime_lims) +
  ylim(c(2, 13)) +
  xlab("Adjusted time (years)") +
  ylab("Biomarker value") +
  ggtitle("Align using last visit") +
  theme_bw(base_size = 20) +
  theme(legend.position = "inside", legend.position.inside = c(.15, .8))

res_estimate_first <- sila_estimate(res$tout, df, align_event = "first")

ggplot(res_estimate_first, aes(x = estdtt0, y = val, color = Subject)) +
  geom_point() +
  geom_line(aes(group = Subject)) +
  geom_line(
    data = res$tout,
    aes(x = adtime, y = val),
    linewidth = 2,
    color = "gray",
    linetype = "dashed"
  ) +
  xlim(adtime_lims) +
  ylim(c(2, 13)) +
  xlab("Adjusted time (years)") +
  ylab("Biomarker value") +
  ggtitle("Align using first visit") +
  theme_bw(base_size = 20) +
  theme(legend.position = "inside", legend.position.inside = c(.15, .8))

res_estimate_all <- sila_estimate(res$tout, df, align_event = "all")

ggplot(res_estimate_all, aes(x = estdtt0, y = val, color = Subject)) +
  geom_point() +
  geom_line(aes(group = Subject)) +
  geom_line(
    data = res$tout,
    aes(x = adtime, y = val),
    linewidth = 2,
    color = "gray",
    linetype = "dashed"
  ) +
  xlim(adtime_lims) +
  ylim(c(2, 13)) +
  xlab("Adjusted time (years)") +
  ylab("Biomarker value") +
  ggtitle("Align using all visits") +
  theme_bw(base_size = 20) +
  theme(legend.position = "inside", legend.position.inside = c(.15, .8))
```

## Development

### Style

#### Git

This repo follows the
[Conventional Commits](https://www.conventionalcommits.org/en/v1.0.0/#summary) 
specification for commit messages.
Pre-commit will automatically check your commit message and fail your commit
if your commit message doesn't conform to the specification.
